Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis

### 2015-06-01 paper/main.tex
clarify testC, hopefully; update docs
last edits before resubmission

% \subsection*{Keywords}
% 
\maketitle

@@

A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:2002}, details some of the benefits and faults of the most popular indices.  According to these reviews, a majority of the indices give comparable results, save Strauss's linear index $L$, despite the fact that most of the methods differ by range and linearity of response.  While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:2002}.  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:2002} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).  The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors and thus hypothesis testing difficult.  This is, in fact, a common fault amongst most of the indices.

\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for formal hypothesis testing.  Much of this work however focuses on general resource selection and doesn't directly address the problems of food selection.  Our model focuses specifically on the problem of predators' food selection, offers formal inference and hypothesis testing similar to that of GLMs \citet{Manly:2002}, and provides meaningful single number summaries of the predators' food preferences.  The model we are proposing 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers formal statistical justification similar to GLMs, 3) accommodates data collected from multiple prey species across multiple time points, and 4) provides parameters estimates even when exact counts of prey species eaten are not fully observed.

% To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.  Outcomes of our model, are that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.

@@

To illustrate these methods, we analysed a data set collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).  Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.  Spiders were removed from the leaf litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.  In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.  For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.  On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.  The range of the sample sizes across all $18$ months was $11$ to $181$ for caught spiders, $7$ to $322$ for trapped Diptera, and $101$ to $755$ for trapped Collembola.  Figure~\ref{fig:trapped} plots the total number of each order that was caught during each time period.

\begin{table}
  \label{tab:s1}
  \centering
  \begin{tabular}{llll}
    \hline
    \textbf{Target group} & \textbf{Primer names and sequences} $5'-3'$ & \textbf{Size (bp)} & \textbf{Source}\\
    \hline
    Collembola & Col3F: GGACGATYTTRTTRGTTCGT & 228 & \citet{Sint:2012} \\
    & Col-gen-A246: TTTCACCTCTAACGTCGCAG & & \\
    Diptera & DIPS16: CACTTGCTTCTTAAATrGACAAATT & 198 & \citet{Eitzinger:2014} \\
    & DIPA17: TTyATGTGAACAGTTTCAGTyCA &  & \\
  \end{tabular}
  \caption{Targeted prey orders, primer names and sequences, size of amplicon, and source of design for the detection of prey taxa within the guts of Schizocosa spiders.  Both primer sets were used in singleplex PCR assays.}
\end{table}

@@

With point estimates of $c_{st}$ under the model $c_{st} = \lambda_{st}/\gamma_{st}$, we can test any number of linear contrasts.  For instance, the hypotheses $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$ state that wolf spiders equally prefer the orders Diptera and Collembola at each of the $18$ time points.  Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.

@@

The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002}, produced one number summaries that rarely considered more than one prey species, did not take into account changes across time, and had minimal statistical justification.  Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.  Further, these indices could not handle unobserved count data.

A more recent method for testing prey preferences was presenteby by Agusti et al (2003).  This method attempts to test the hypothesis that individuals are eating prey at random using a test based on Monte Carlo simulation (more specificially the bootstrap).  Unfortunately, we believe that the method described is flawed specifically because it does not attempt to model the distribution of the number of prey items eaten, as we have done.  The null hypothesis tested by Agusti et all (2003) is that the proportion of prey species $j=1,\ldots,J$ in the mesocosm, denoted $\pi_j$, is equal to the proportion of prey species $j$ in the predator's guts, denoted by $p_j$, at all of the sampling times.  This is tested by repeatedly simulating the gut contents of spiders in each sample under this hypothesis and then comparing summary statistics computed from the observed data and the simulated data to generate a $p$-value.  The problem with this approach is that the relationship between $\pi_j$ and $p_j$ depends on the number of prey items eaten by each predator.  If each predator had eaten exactly one prey item then the null hypothesis of Agusti et al (2003) would be equivalent to random sampling of prey.  However, the relationship between $\pi_j$ and $p_j$ is more complicated when predators can have more than one prey item in their gut.  Given that a predator has $m$ prey items in its gut, randomly sampled from the mesocosm, the marginal probability that at least one item from prey species $j$ is present is $1-(1-\pi_j)^m$.  The marginal probability that an individual has as least one item from prey species $j$ in its gut is then $\sum_{m=1}^{\infty} (1-(1-\pi_j)^m)P(m)$, where $P(m)$ is the probability that a predator has a total of $m$ species in its gut.  The presence of species in a predators gut contents also cannot be simulated without modeling the number of items in the gut for exactly the same reason.  This shows the importance of the distribution of prey items in the gut in testing the hypothesis of random sampling and because of this it is unclear what hypothesis is being tested by Agusti et al (2003).  Monte Carlo simulation methods similar to this could be implemented using the output from our model, but these would be computationally intensive in comparison to the likelihood ratio tests which can be computed almost immediately.

Throughout this paper we assume the complete data were generated from underlying Poisson distributions.  Though this assumption is not easily tested, and other factors affecting our model may distort our methods, e.g.\ sampling issues, environmental variables, differential decay rates, we do not believe a better assumption is readily available to handle all the cases we considered.  The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.  One could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.  Such a distribution could theoretically perform better than our model when full count data are observed and the sampled data were not generated from a Poisson distribution -- if the data were generated from a Poisson distribution then our likelihood ratio test ensures we are using the most powerful test available.  But when only unobserved count data is available, any distribution with more than one parameter is unidentifiable.  The Poisson distribution on the other hand allowed us to use a relatively simple EM algorithm to estimate parameters of interest when only unobserved count data are available.  Future development of our methods might consider more complex distributions to handle variables not presently considered in the case of fully observed count data.

The methods we have presented offer ecologists a formal statistical framework to model and test predators feeding preferences across an array of time points and prey species. Building from intuitive assumptions, we have developed methodologies for analyzing counts of prey items consumed and have extended these methods for analyzing simple presence/absence data. We have provided computationally efficient methods for fitting these models using the EM algorithm and developed powerful tests for a hierarchy of hypotheses based on the likelihood ratio. We are not aware of previous methods that have offered such abilities. These methods are available to researchers via the R package spiders which is available at \url{http://cran.r-project.org/web/packages/spiders/index.html}.

>>>>

\maketitle

@@

A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:2002}, details some of the benefits and faults of the most popular indices.  According to these reviews, a majority of the indices give comparable results, save the linear index $L$ by \citet{Strauss:1979}, despite the fact that most of the methods differ by range and linearity of response.  While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:2002}.  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:2002} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).  The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors and thus hypothesis testing difficult.  This is, in fact, a common fault amongst most of the indices.

\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for formal hypothesis testing.  Much of this work focuses on general resource selection and doesn't directly address the problems of food selection.  By combining the food selectivity literature with formal statistical modelling, our new method 1) focuses specifically on the problem of predators' food selection, while maintaining the intuitiveness of the indices summarised by \citet{Lechowicz:1982} and \citet{Manly:2002}, 2) offers formal statistical justification, enabling proper inference and hypothesis testing similar to that of GLMs \citep{Manly:2002}, 3) accommodates data collected from multiple prey species across multiple time points, 4) provides parameter estimates even when exact counts of prey species eaten are not fully observed, and 5) provides meaningful single number summaries of the predators' food preferences.

@@

To illustrate these methods, we analysed a data set collected to investigate the feeding preferences of two species of wolf spider, \textit{Schizocosa ocreata} and \textit{Schizocosa stridulans} (Araneae: Lycosidae).  Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.  Spiders were removed from the leaf litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% EtOH, and preserved at $-20\celsius$ until DNA extraction.  In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.  For the analysis, both species of \textit{Schizocosa} were pooled and the number of spiders and prey were analysed by month.  On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.  The range of the sample sizes across all $18$ months was $11$ to $181$ for caught spiders, $7$ to $322$ for trapped Diptera, and $101$ to $755$ for trapped Collembola.  Figure~\ref{fig:trapped} plots the total number of each order that was caught during each time period.

\begin{table}
  \label{tab:s1}
  \centering
  \begin{tabular}{llll}
    \hline
    \textbf{Target group} & \textbf{Primer names and sequences} $5'-3'$ & \textbf{Size (bp)} & \textbf{Source}\\
    \hline
    Collembola & Col3F: GGACGATYTTRTTRGTTCGT & 228 & \citet{Sint:2012} \\
    & Col-gen-A246: TTTCACCTCTAACGTCGCAG & & \\
    Diptera & DIPS16: CACTTGCTTCTTAAATrGACAAATT & 198 & \citet{Eitzinger:2014} \\
    & DIPA17: TTyATGTGAACAGTTTCAGTyCA &  & \\
  \end{tabular}
  \caption{Targeted prey orders, primer names and sequences, size of amplicon, and source of design for the detection of prey taxa within the guts of \textit{Schizocosa} spiders.  Both primer sets were used in singleplex PCR assays.}
\end{table}

@@

With point estimates of $c_{st}$ under the model $c_{st} = \lambda_{st}/\gamma_{st}$, we can test any number of linear contrasts.  For instance, the hypotheses $c_{1t} = c_{2t}$, for $t \in \{1, \ldots, 18\}$ state that wolf spiders equally prefer the orders Diptera and Collembola at each of the $18$ time points.  Using a level of significance of $0.05$, and after making a Bonferroni multiple comparisons adjustment, the data can not say that the two prey are differently preferred in October, November, and December of $2011$ and for March and July of $2012$.  In all other time periods, the data suggest that Diptera and Collembola are unequally preferred by wolf spiders.  Contrasts such as these allow molecular ecologists to identify in greater detail and formally test changes in prey preferences across time points and multiple species.

@@

The earliest estimators of predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002}, produced one number summaries that rarely considered more than one prey species, did not take into account changes across time, and had minimal statistical justification.  Instead, they were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.  Further, these indices could not handle unobserved count data.

A more recent method for testing prey preferences was developed by Agusti et al (2003).  This method attempts to test the hypothesis that individuals are eating prey at random using a test based on Monte Carlo simulation (more specificially the bootstrap).  Unfortunately, the method described does not attempt to model the distribution of the number of prey items eaten, as we have done, and therefore we provide a methodological advance on the approach described previously.  The null hypothesis tested by Agusti et all (2003) is that the proportion of prey species $j=1,\ldots,J$ in the mesocosm, denoted $\pi_j$, is equal to the proportion of prey species $j$ in the predator's guts, denoted by $p_j$, at all of the sampling times.  This is tested by repeatedly simulating the gut contents of spiders in each sample under this hypothesis and then comparing summary statistics computed from the observed data and the simulated data to generate a $p$-value.  The problem with this approach is that the relationship between $\pi_j$ and $p_j$ depends on the number of prey items eaten by each predator.  If each predator had eaten exactly one prey item then the null hypothesis of Agusti et al (2003) would be equivalent to random sampling of prey.  However, the relationship between $\pi_j$ and $p_j$ is more complicated when predators can have more than one prey item in their gut.  Given that a predator has $m$ prey items in its gut, randomly sampled from the mesocosm, the marginal probability that at least one item from prey species $j$ is present is $1-(1-\pi_j)^m$.  The marginal probability that an individual has as least one item from prey species $j$ in its gut is then $\sum_{m=1}^{\infty} (1-(1-\pi_j)^m)P(m)$, where $P(m)$ is the probability that a predator has a total of $m$ species in its gut.  The presence of species in a predator's gut contents also cannot be simulated without modeling the number of items in the gut for exactly the same reason.  This demonstrates the importance of knowing the distribution of prey items in the gut in testing the hypothesis of random sampling.  Monte Carlo simulation methods similar to this could be implemented using the output from our model, but these would be computationally intensive in comparison to the likelihood ratio tests which can be computed almost immediately.

Throughout this paper we assume the complete data were generated from underlying Poisson distributions.  Though this assumption is not easily tested, and other factors affecting our model may distort our methods (e.g.\ sampling issues, environmental variables, differential decay rates), we do not believe a better assumption is readily available to handle all the cases we considered.  The Poisson distribution provides reasonable biological interpretation to the available data, data obtained by ecologists interested in prey electivity.  One could hypothesize a more complex distribution that similarly has good biological interpretation, say a distribution with more than one parameter.  Such a distribution could theoretically perform better than our model when full count data are observed and the sampled data were not generated from a Poisson distribution -- if the data were generated from a Poisson distribution then our likelihood ratio test ensures we are using the most powerful test available.  However, when only unobserved count data is available, any distribution with more than one parameter is unidentifiable.  The Poisson distribution on the other hand allowed us to use a relatively simple EM algorithm to estimate parameters of interest when only unobserved count data are available.  Future development of our methods might consider more complex distributions to handle variables not presently considered in the case of fully observed count data.

The methods we have presented offer ecologists a formal statistical framework to model and test predator feeding preferences across an array of time points and prey species. Building from intuitive assumptions, we have developed methodologies for analyzing counts of prey items consumed and have extended these methods for analyzing simple presence/absence data. We have provided computationally efficient methods for fitting these models using the EM algorithm and developed powerful tests for a hierarchy of hypotheses based on the likelihood ratio. We are not aware of previous methods that have offered such abilities. These methods are available to researchers via the R package spiders which is available at \url{http://cran.r-project.org/web/packages/spiders/index.html}.
